Understanding the Impact of Mortgage State and Income on Household Poverty Status

### 2024-04-18 paper/US_Poverty_Analysis_2019.qmd
edit grammar

\newpage
# Introduction
Despite many policies the United States have implemented to mitigate poverty over the past decades, there has been no real improvement [@Kearney_Harris]. Poverty remains a major issue in the US, affecting millions of people. In 2019, the poverty rate was 10.5% [@Semega_2022]. Therefore, it is important to find out the causes of poverty, in order to reduce it. Multiple research, using the Official Poverty Measure, was conducted to look at why poverty persists. However, the Official Poverty Measure is flawed in various aspects, such as not accounting for regional differences in costs of living and government benefits  [@Gesumaria_2015]. So we will analyze why some US household are in poverty in 2019 using the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset from the United States Census Bureau (USCB) [@Bureau_2022]. USCB is an agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy [@uscb]. Supplemental Poverty Measure was developed to address numerous concerns of Official Poverty Measure critics, providing an improved statistical picture of poverty [@Gesumaria_2015]. We build a prediction model for households poverty status using 2019 Revised SPM dataset. We primarily look at two factors: mortgage status and income in each household. This allows us to look at the question whether potential financial factors may influence the poverty status for US households in 2019.

In this paper, a logistic regression model is used to predict the poverty status of households in the United States in 2019, with data from the 2019 Revised Supplemental Poverty Measure Research dataset. Logistic regression is a great choice since it is used to predict binary outcomes, such as poverty status (in poverty or not in poverty). Our analysis focuses on estimating the likelihood of household being in poverty, based on various financial factors captured in the SPM dataset. We select 17 data features for analysis. After some data cleaning, we summarize 17 features into 2 variables: 1) mortgage state and 2) household annual total income after tax, plus government benefits and minus expenses.

@@

## Data Source
This report uses the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset provided by the United States Census Bureau as our main source of data. The Census Bureau provides information about the people and economy of the United States, with a goal to support economic growth, enhance scientific knowledge, and assist in making informed decisions. The official poverty measure has been criticized due to its multiple limitations [@sgp2019]. For example, it does not consider benefits from most of the largest programs that aid the low-income population. For instance, it uses money income before taxes, so it does not measure the income available for individuals to spend, which for most people is after-tax income. Therefore, any effects of tax credits designed to assist persons with low income are not considered by the official measure. The Supplemental Poverty Measure (SPM), similar to the official poverty measure, tackles the shortcomings of poverty assessment. It determines poverty status by gauging resources against a defined standard of living. This standard considers expenditures on essentials like food, clothing, shelter, and utilities, plus a bit more for additional expenses. The resources assessed include disposable income, accounting for taxes and some non-cash benefits, available to cover these needs [@Gesumaria_2015]. In 2019, the dataset contained a total of 157,959 entries.

>>>>

\newpage
# Introduction
Despite many policies the United States have implemented to mitigate poverty over the past decades, there has been no real improvement [@Kearney_Harris]. Poverty remains a major issue in the US, affecting millions of people. In 2019, the poverty rate was 10.5%, which was approximately 34.0 million people [@Semega_2022]. Therefore, we want to find out the causes of poverty, in order to reduce it. Multiple research, using the Official Poverty Measure dataset gathered by the United States Census Bureau (USCB), looked at why poverty persists. USCB is an agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy  [@uscb]. Over the years, people have realized that the Official Poverty Measure is flawed in various aspects, such as not accounting for regional differences in costs of living, not considering household's mortgage state, and government benefits programs  [@Gesumaria_2015]. Therefore, Supplemental Poverty Measure was developed to address numerous concerns of Official Poverty Measure critics, providing an improved statistical picture of poverty . In this paper, we will analyze why some US household are in poverty in 2019 using the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset from the United States Census Bureau (USCB) [@Bureau_2022]. We build a prediction model for households poverty status using 2019 Revised SPM dataset. This allows us to look at the question whether potential financial factors may influence the poverty status for US households in 2019.

In this paper, a logistic regression model is used to predict the poverty status of households in the United States in 2019, with data from the 2019 Revised Supplemental Poverty Measure Research dataset. Logistic regression is a great choice since it is used to predict binary outcomes, such as poverty status (in poverty or not in poverty). Our analysis focuses on estimating the likelihood of household being in poverty, based on various financial factors captured in the SPM dataset. I selected 17 data features for analysis. After some data cleaning, I summarized 17 features into 2 variables: 1) mortgage state and 2) household annual total income after tax, plus government benefits and minus expenses.

@@

## Data Source
This report uses the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset provided by the United States Census Bureau as our main source of data. The Census Bureau provides information about the people and economy of the United States, with a goal to support economic growth, enhance scientific knowledge, and assist in making informed decisions. The official poverty measure has been criticized due to its multiple limitations [@sgp2019]. For example, it does not consider benefits from most of the largest programs that aid the low-income population. For instance, it uses money income before taxes, so it does not measure the income available for individuals to spend, which for most people is after-tax income. Therefore, any effects of tax credits designed to assist persons with low income are not considered by the official measure. So, the Supplemental Poverty Measure (SPM) was developed. It is similar to the official poverty measure, but it is more comprehensive. It considers expenditures on essentials like food, clothing, shelter, and utilities, plus a bit more for additional expenses. The resources assessed include disposable income, accounting for taxes and some non-cash benefits, available to cover these needs [@Gesumaria_2015]. In 2019, the dataset contained a total of 157,959 entries.
